Exploring probability distributions for bivariate temporal granularities

### 2019-11-24 main.Rmd
visualization intro delete

   Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within the tidy workflow, so that probability distributions can be examined using the range of graphics available in [@Wickham2009-pk]. 
bibliography: bibliography.bib
output:
  bookdown::pdf_book:
    #base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
---

```{r initial, echo = FALSE, cache = FALSE, include = FALSE}
options("knitr.graphics.auto_pdf" = TRUE)
library(knitr)
library(tidyverse)
library(lubridate)
library(lvplot)
library(ggridges)
library(viridis)
library(tsibble)
library(gravitas)
library(tsibble)
opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, comment = "#>",
  fig.path = "figure/", fig.align = "center", fig.show = "hold",
  cache = FALSE, cache.path = "cache/",
  out.width = ifelse(is_html_output(), "100%", "\\textwidth")
)
knitr::opts_knit$set(root.dir = here::here())
```

@@

\begin{equation}\label{eq6}
A_{BG, AG}(z) = BG(z - \sum_{w=0}^{k-1}\vert T_{w \mod R'} \vert) for \z \in T_k
\end{equation}
where,  
z denotes the index set (index of BG),
R' denotes the number of granules of AG in each period of BG,  
$T_w$ are the sets of indices of BG describing $AG_w$ such that $AG(w) = \bigcup_{z \in T_w}BG(z)$, and $T_w = 0$ for $w \epsilon Z_{<0}$  
$\vert T_w \vert$ is the cardinality of set $T_w$.

@@

\begin{equation} \label{eq4}
\begin{split}
A(BG,AG)(8) & = BG(8 - \sum_{w=0}^{3-1}\vert T_{w \mod 2}\vert) , since \8 \in T_{3} \\
  & = BG(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\
  & = BG(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\
  & = BG(8 - \vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert - \vert T_{2 \mod 2}\vert) \\
  & = BG(8 - 2*\vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert) \\
  & = BG(8 - 2.2 - 3) \\
  & = BG(1) \\
\end{split}
\end{equation}

\begin{equation} \label{eq5}
\begin{split}
A(BG,AH)(10) & = BG(10 - \sum_{w=0}^{1-1}\vert T_{w \mod 2}\vert) , since \10 \in T_{1}  \\
  & = BG(10 - \sum_{w=0}^{0}\vert T_{w\mod 2}\vert) \\
  & = BG(10 - \vert T_{0 \mod 2}\vert) \\
  & = BG(10 - \vert T_{0}\vert) \\
  & = BG(10 - 7) \\
  & = BG(3) \\
\end{split}
\end{equation}

@@

### Computation of multiple-order from single-order

#### All single order granularities are circular

\begin{equation} \label{eq3}
\begin{split}
C_{(G_l,G_m)}(z) & = C_{(G_l,G_(l+1)}(z) + k(l, l+1)(C_{(G_(l+1),G_m)}(z)-1) \\  
  & =   C_{(G_l,G_(l+1)}(z) + k(l, l+1)[ C_{G_(l+1),G_(l+2)}(z) + k(l+1, l+2)( C_{G_(l+2), G_m)}(z) - 1) - 1]\\
  & =  C_{(G_l,G_(l+1)}(z) + k(l, l+1)(C_{G_(l+1),G_(l+2)}(z) - 1) + k(l, l+1)k(l+1, l+2)(C(l+2, m) - 1)\\
  & =  C_{(G_l,G_(l+1)}(z) + k(l, l+1)(C_{G_(l+1),G_(l+2)}(z) - 1) + k(l, l+2)(C_{G_(l+2),G_(l+m)}(z) - 1)\\
  &\vdots\\
  & = \sum_{i=0}^{order(m) - order(l) - 1} k(l, l+i)(C_{G_(l+i),G_(l+i+1)}(z) - 1)\\
\end{split}
\end{equation}

@@

#### Single order granularities are circular or aperiodic

#### All single order granularities are circular

### Computation of single-order from multiple-order

@@

<!-- In the next section, we discuss the computation of any single-order-up and multiple-order-up granularities in a periodic and aperiodic set up. -->

#### Single order granularities are circular or aperiodic

<!-- Computation of time granularities would differ in a periodic and non-periodic set up. -->

<!-- Now, let us discuss the computation of circular granularities. -->

<!-- #### Aperiodic single-order-up granularities -->

<!-- Aperiodic single-order-up granularities refers to the granularities which are formed with units that does not repeat its values in regular intervals or periods. In Gregorian calendar, examples may include days of the month where each month may consists of 28, 29, 30 or 31 days. So there is no single number that can be used for converting days to months. We cannot compute the aperiodic single-order-up granularities using the index of the tsibble and modular arithmetic like in the periodic case using just the hierarchy table and relationships of units. -->

<!-- ### Multiple order-up granularities  -->

<!-- Computation of multiple-order-up granularities will differ if the units in the hierarchy table are periodic or aperiodic. We split the method of computation into two cases - one concerning all periodic single-order-up granularities and the second with mixed single-order-up granularities. -->

<!-- ### All single-order-up granularities are periodic -->

<!-- z is the index set of the tsibble and x, y are two linear granularities with  $order(x) <  order(y)$. Also, $f(x, y)$ denotes the accessor function for computing circular granularity which relates x and y and $c(x, y)$ is a constant which relates x and y. It is easy to see that for $order(x+1) =  order(y)$, the function is same as the single-order-up granularities. -->

<!-- Then, the accessor function f can be used recursively to obtain any multiple-order-up granularities as follows: -->

<!-- \begin{equation} \label{eq1} -->
<!-- \begin{split} -->
<!-- f(x,y) & = f(x, x+1) + c(x, x+1)(f(x+1, y)-1) \\   -->
<!--   & =  f(x, x+1) + c(x, x+1)[f(x+1, x+2) + c(x+1, x+2)(f(x+2, y) - 1) - 1]\\ -->
<!--   & =  f(x, x+1) + c(x, x+1)(f(x+1, x+2) - 1) + c(x, x+1)c(x+1, x+2)(f(x+2, y) - 1)\\ -->
<!--   & =  f(x, x+1) + c(x, x+1)(f(x+1, x+2) - 1) + c(x, x+2)(f(x+2, y) - 1)\\ -->
<!--   &\vdots\\ -->
<!--   & = \sum_{i=0}^{order(y) - order(x) - 1} c(x, x+i)(f(x -->
<!--   +i, x+i+1) - 1)\\ -->
<!-- \end{split}x -->
<!-- \end{equation} -->

<!-- Let us use the equation to compute the multiple-order-up granularity uinal_katun for Mayan calendar, which is periodic. -->

<!-- From Equation(1), we have -->
<!-- \begin{equation} \label{eq1} -->
<!-- \begin{split} -->
<!-- f(uinal, baktun) & = f(uinal, tun) + c(uinal, tun)f(tun,katun) + c(uinal, katun)f(katun, baktun) \\ -->
<!--               & = \lfloor z/20\rfloor \mod 18  + 20*\lfloor z/20*18\rfloor \mod 20 + 20*18*20\lfloor z/20*18*20\rfloor \mod 20 \\ -->
<!-- \end{split} -->
<!-- \end{equation} -->
<!--              (using equations \ref{single_order}) -->

<!-- ### Mixed single-order-up granularities - circular or aperiodic -->

<!-- Let us turn to Gregorian calendar for addressing this case. -->

<!-- Suppose we have a hierarchy table for  Gregorian calendar as \ref{table_accessor2}. Since months consists of unequal number of days, any temporal unit which is higher in order than months will also have unequal number of days. This is an example of a hierarchy table which has both periodic and aperiodic single-order-up granularities. The single-order-up granularity day_month is aperiodic. Any single-order-up granularities which are formed by units below days are periodic. Similarly, all single-order-up granularities which are formed using units whose orders are higher than months are also periodic. -->

<!-- <!-- Hence, any calendar categorization which includes one linear granularity with orders at most as days and another one with orders at least as months will be aperiodic in nature. However, it is very much possible for a calendar categorization consisting of two linear granularities higher in order than month to be periodic.For example, the categorization month_quarter would be periodic since each quarter always consists of three months. --> -->

<!-- ```{r table_accessor2} -->
<!-- kable(tibble::tibble( -->
<!--   `linear granularities` = c("minute", "hour", "day", "month", "quarter", "year"), -->
<!--   `Conversion factor` = c(60,24, "aperiodic", 3, 4, 1), format = "markdown")) -->
<!-- ``` -->

<!-- There can be three scenarios for obtaining calendar categorization here: -->
<!-- - granularities consisting of two units whose orders are less than day   -->
<!-- - granularities consisting of two units whose orders are more than month   -->
<!-- - granularities consisting of one unit with order at most day and another with order at least month   -->

<!-- The calendar categorization resulting from the first two cases are periodic and has been handled in the earlier section. -->

<!-- The calendar categorization resulting from the last case are aperiodic. Examples might include day of the quarter or hour of the month. In this section, we will see how to obtain aperiodic circular granularities of these types. -->

<!-- \begin{equation} \label{eq1} -->
<!-- \begin{split} -->
<!-- f_{(hour, month)}(z) = f_{(hour, day)}(z) + c(hour, day)f_{(day, month)}(z)\\ -->
<!-- \end{split} -->
<!-- \end{equation} -->

<!-- Here, the first part of the equation is a single-order-up granularity which can be obtained using last section. The second part is not single-order-up and can not be broken down further since each month consists of different number of days. In this case, it is important for us to know which month of the year and if the year is a leap year to obtain day of the month. -->

@@

The grammar of graphics introduced a framework to incorporate underlying features for statistical graphics to construct them by relating data space to the graphic space [@Wilkinson1999-nk]. The layered grammar of graphics proposed by [@Wickham2009-pk], which is a modification over the former, allows us to produce graphics using the same structured thinking that is used to design analysis. Briefly, any graphics are made up of **Layers**, **Facet**,**Scale** and a  **coordinate system**. **Layers** are made up of **data**, **mapping**, **statistical transformation**, **geometric objects** and **position adjustments**.

Creating a visualisation requires a number of nuanced judgements. However, we exploit some features of layers and faceting to come up with a recommendation system while visualizing the distribution of the "dependent" variable across bivariate temporal granularities. The general framework of visualizing the distribution invloves a faceting approach with one temporal granularity plotted along the x-axis, the other one across facet and the dependent time series variable on the y-axis. Different distribution plots might be appropriate depending on which features of the distrbution we are interested to look at, the levels of time granularities being plotted, how granularities interact and number of observations available. We will look at each of these aspects separately and analyze the effect of each on visualization.

@@

@@

<!-- ```{r write-bib} -->
<!-- write_bib(c("ggplot2", "tidyverse", "sugrrants", "geofacet", "shiny", "wanderer4melb"), 'packages.bib') -->
<!-- ``` -->

>>>>

   Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within the tidy workflow, so that probability distributions can be examined using the range of graphics available in [@Wickham2009-pk]. 
bibliography: [bibliography.bib]
output:
  bookdown::pdf_book:
    #base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
---

```{r initial, echo = FALSE, cache = FALSE, include = FALSE}
options("knitr.graphics.auto_pdf" = TRUE)
library(knitr)
library(tidyverse)
library(lubridate)
library(lvplot)
library(ggridges)
library(tsibble)
library(gravitas)
library(tsibble)
opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, comment = "#>",
  fig.path = "figure/", fig.align = "center", fig.show = "hold",
  cache = FALSE, cache.path = "cache/",
  out.width = ifelse(is_html_output(), "100%", "\\textwidth")
)
knitr::opts_knit$set(root.dir = here::here())
```

@@

\begin{equation}\label{eq6}
A_{BG, AG}(z) = BG(z - \sum_{w=0}^{k-1}\vert T_{w \mod R'} \vert) for z \in T_k
\end{equation}
where,  
z denotes the index set (index of BG),
R' denotes the number of granules of AG in each period of BG,  
$T_w$ are the sets of indices of BG describing $AG_w$ such that $AG(w) = \bigcup_{z \in T_w}BG(z)$, and $T_w = 0$ for $w \epsilon Z_{<0}$  
$\vert T_w \vert$ is the cardinality of set $T_w$.

@@

\begin{equation} \label{eq4}
\begin{split}
A(BG,AG)(8) & = BG(8 - \sum_{w=0}^{3-1}\vert T_{w \mod 2}\vert) , since 8 \in T_{3} \\
  & = BG(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\
  & = BG(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\
  & = BG(8 - \vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert - \vert T_{2 \mod 2}\vert) \\
  & = BG(8 - 2*\vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert) \\
  & = BG(8 - 2.2 - 3) \\
  & = BG(1) \\
\end{split}
\end{equation}

\begin{equation} \label{eq5}
\begin{split}
A(BG,AH)(10) & = BG(10 - \sum_{w=0}^{1-1}\vert T_{w \mod 2}\vert) , since 10 \in T_{1}  \\
  & = BG(10 - \sum_{w=0}^{0}\vert T_{w\mod 2}\vert) \\
  & = BG(10 - \vert T_{0 \mod 2}\vert) \\
  & = BG(10 - \vert T_{0}\vert) \\
  & = BG(10 - 7) \\
  & = BG(3) \\
\end{split}
\end{equation}

@@

### Computation of multiple-order from single-order {#sec:multiplefromsingle}

\begin{equation} \label{eq3}
\begin{split}
C_{(G_l,G_m)}(z) & = C_{(G_l,G_(l+1)}(z) + k(l, l+1)(C_{(G_(l+1),G_m)}(z)-1) \\  
  & =   C_{(G_l,G_(l+1)}(z) + k(l, l+1)[ C_{G_(l+1),G_(l+2)}(z) + k(l+1, l+2)( C_{G_(l+2), G_m)}(z) - 1) - 1]\\
  & =  C_{(G_l,G_(l+1)}(z) + k(l, l+1)(C_{G_(l+1),G_(l+2)}(z) - 1) + k(l, l+1)k(l+1, l+2)(C(l+2, m) - 1)\\
  & =  C_{(G_l,G_(l+1)}(z) + k(l, l+1)(C_{G_(l+1),G_(l+2)}(z) - 1) + k(l, l+2)(C_{G_(l+2),G_(l+m)}(z) - 1)\\
  &\vdots\\
  & = \sum_{i=0}^{m - l - 1} k(l, l+i)(C_{G_(l+i),G_(l+i+1)}(z) - 1)\\
\end{split}
\end{equation}

@@

### Computation of single-order from multiple-order {#sec:singlefrommultiple}

@@

In \ref{sec:singlefrommultiple} and \ref{sec:multiplefromsingle}, we assume that all single order granularities are circular.What happens when we have a mix of circular and aperiodic granularities.

Let us turn to Gregorian calendar for addressing this case. Suppose we have a hierarchy table for  Gregorian calendar as \ref{table_accessor2}. Since months consists of unequal number of days, any temporal unit which is higher in order than months will also have unequal number of days. This is an example of a hierarchy structure which has both circular and aperiodic single-order-up granularities. The single-order-up granularity day_month is aperiodic. Any single-order-up granularities which are formed by units below days are periodic. Similarly, all single-order-up granularities which are formed using units whose orders are higher than months are also periodic.

<!-- Hence, any calendar categorization which includes one linear granularity with orders at most as days and another one with orders at least as months will be aperiodic in nature. However, it is very much possible for a calendar categorization consisting of two linear granularities higher in order than month to be periodic.For example, the categorization month_quarter would be periodic since each quarter always consists of three months. -->

|  G	        |C/A               |k
|------------	|------------------|------------|
|  minute     |minute-of-hour    |  60        |
|  hour    	  |hour-of-day       |  24  	    |
|  day      	|day-of-month      | "aperiodic"|
|  month    	|month-of-year     |  12        |
|  year   	  | 1                |	1         |

There can be three scenarios for obtaining multiple-order-granularities here:
- granularities relating two linear granularities whose orders are less than day
- granularities relating two linear granularities whose orders are more than month
- granularities relating two linear granularities with order at most day and another with order at least month

The single order-up granularites resulting from the first two cases are periodic and has been handled in \ref{sec:singlefrommultiple} and \ref{sec:multiplefromsingle}. The calendar categorization resulting from the last case are aperiodic. Examples might include day of the quarter or hour of the month. Let us consider the computation of $A_{hour, month}$ given single order-up granularities $A_{hour, day} = 12$, $A_{day, month} = 20$ and $C_{month, year} = 9$.

\begin{equation} \label{eq1}
\begin{split}
A_{(hour, month)} & = (\sum_{w=0}^{9 - 1}(\vert T(w)\vert) +  20)/24 + 12 \\
                  & = (\sum_{w=0}^{C_{month, year} - 1}(\vert T(w)\vert) +  A_{day, month})*k(hour_day) + A_{hour, day} \\
                  & = (A{day_year} - 1)*k(hour_day) + A_{hour, day} \\
\end{split}
\end{equation}
where, $month(w) = \bigcup_{z \in T_w}day(z)$

Clearly, the computation of multiple order-up granularities become more involved with aperiodic mappings of linear granularities in the calendar structure. With one aperiodic granularity, we can write the following:

A_{P, Q)} & = (A{AG_Q} - 1)*k(P_AG) + A_{P, AG} {TO REVISE: many aperiodic granularties not to be included}

@@

<!-- The grammar of graphics introduced a framework to incorporate underlying features for statistical graphics to construct them by relating data space to the graphic space [@Wilkinson1999-nk]. The layered grammar of graphics proposed by [@Wickham2009-pk], which is a modification over the former, allows us to produce graphics using the same structured thinking that is used to design analysis. Briefly, any graphics are made up of **Layers**, **Facet**,**Scale** and a  **coordinate system**. **Layers** are made up of **data**, **mapping**, **statistical transformation**, **geometric objects** and **position adjustments**. -->

Creating a visualisation requires a number of nuanced judgements. We exploit some features of layers and faceting [@Wilkinson1999-nk] and [@Wickham2009-pk] to come up with a recommendation system while visualizing the distribution of the "dependent" variable across bivariate temporal granularities. The general framework of visualizing the distribution invloves a faceting approach with one temporal granularity plotted along the x-axis, the other one across facet and the dependent time series variable on the y-axis. Different distribution plots might be appropriate depending on which features of the distrbution we are interested to look at, the levels of time granularities being plotted, how granularities interact and number of observations available. We will look at each of these aspects separately and analyze the effect of each on visualization.

@@

```{r write-bib}
write_bib(c("ggplot2", "tidyverse", "sugrrants", "geofacet", "shiny", "wanderer4melb"), 'packages.bib')
```

@@
